Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2013-10-26 omrp.tex
Streamlining Sec 4.4 and 4.5
format changes + a bit more explanations sprinkled here and there

Meketon and Schmeiser \citep{Meketon1984} show that $\frac{n_l}{m_l}\var{n_l VD_l} \rightarrow \alpha \sigma^4_{\xh}$ where $\alpha = \frac{4}{3}$ for maximally overlapping batches and $\alpha = 2$ for nonoverlapping batches, resulting in the decrease of variance for the maximally overlapping batches estimator.  
Welch \citep{Welch1987} expands this result to variable amounts of overlap, showing that $\alpha$ is twice the value of (\ref{eq:var_reduct_formula}).  
Damerdji \citep{damerdji1995mean} shows that convergence of $\frac{n_l}{m_l}\var{n_l VD_l}$ to the appropriate $\alpha \sigma^4_{\xh}$, and thus the mean-square consistency of $n_l VD_l$, is guaranteed for $\tfrac{1}{2} + \tfrac{1}{4+\epsilon} < r < 1$ for both nonoverlapping and maximally overlapping batches under the batching structure $m_l = \lfloor n_l^r \rfloor$ and $4+\epsilon$ finite moments assumption A2$^\prime$.  
(With A2, the range of $r$ that ensures mean-square consistency can be found by setting $\epsilon=0$.)

To prove the variance reduction results in our setting, we wish to show that $({n_l}/{m_l}) \left( \var{n_l VG_l} - \var{n_l VD_l} \right) \rightarrow 0$, for which it suffices to show $\sqrt{{n_l}/{m_l}} \left( n_l VG_l - n_l VD_l \right) \xrightarrow{L^2} 0$, where $\xrightarrow{L^2}$ denotes mean-square convergence.

@@

\begin{proof}
	For simplicity of exposition, we drop the subscript $l$ and the $\frac{n_l}{m_l}$ for now and add them in later.  
        Further, we use abbreviations $n_b = \nbl$ and $df = \left(\tfrac{n_l}{m_l} - 1\right) \left(\nbl\right)$.
	\begin{align}
		\e{(nVG - nVD)^2} & = \left( \frac{n}{df} \right)^2 \e{ \left( \sum_{j=1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right] \right)^2 } \notag \\
		& \leq \left( \frac{n}{df} \right)^2 \e{ n_b \sum_{j=1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right]^2 } \label{eq:vvar_conv_pf_1}
	\end{align}
	by the Cauchy-Schwartz inequality.  
        By examining the batch structure in Figure \ref{fig:overlap_nonint} and the definition of $\gbb$ in (\ref{eq:gbb}) one can see that the terms of the expectation are not identically distributed.  
        We will separate these into identically distributed ``center'' batches and ``fringe'' batches.  
        In Figure \ref{fig:overlap_nonint}, only batch 4 is a center batch, while the others are fringe batches.  
        Notice in Figure \ref{fig:overlap_nonint}, we see $\left\lceil \gammab^{-1} \right\rceil = 3$ sets of nonoverlapping batches, each of which has two fringe batches.  
        In general, there are $\left\lceil \gammab^{-1} \right\rceil$ sets of nonoverlapping batches, each of which will have two fringe batches, and $n_b - 2\left\lceil \gammab^{-1} \right\rceil$ center batches.

	The fringe batches will come in identically distributed pairs.  
        The first and last batch would be identically distributed, as would the second and second-to-last, etc.  
        We know this because we could reverse the order of the i.i.d.\ sample and perform the batching again to get the exact same batch structure.  
        Rewriting to take this into account, letting $A_j = (\gb_j - \gbb)^2 - (\db_j - \dbb)^2$, and continuing from (\ref{eq:vvar_conv_pf_1}),
	\begin{align*}
		\left( \frac{n}{df} \right)^2 n_b \e{ \sum_{1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right]^2 } & = \left( \frac{n}{df} \right)^2 n_b \e{\left[ \sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \left[ A_j \right]^2 \
		 + \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \left[ A_j \right]^2 \
		 + \sum_{j=n_b-\left\lceil \gammab^{-1} \right\rceil+1}^{n_b} \left[ A_j \right]^2 \right]\
		} \\
		& = \left( \frac{n}{df} \right)^2 n_b \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 } + \left( \frac{n}{df} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 } \\
		& = \left( \frac{n}{df} \right)^2 n_b \left(n_b - 2\left\lceil \gammab^{-1} \right\rceil \right) \e{ \left[ A_j \right]^2 }
		+ \left( \frac{n}{df} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 }.
	\end{align*}	
	We have $\left\lceil \gammab^{-1} \right\rceil + 1$ expectation terms, each of which has a coefficient that is less than $\left(\frac{n}{df}\right)^2n_b^2 < (2m)^2$.  
        So each expectation can be bounded by
	\begin{align}
		\left( 2m \right)^2 \e{ \left[ A_j \right]^2 } & = \left( 2m \right)^2 \e{ \left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^2 \left[(\gb_j - \gbb) + (\db_j - \dbb)\right]^2 } \notag \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}} \sqrt{\e{\left[(\gb_j - \gbb) + (\db_j - \dbb)\right]^4 }} \label{eq:var_reduct_cauchy2} \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}} \sqrt{8 \left(\e{(\gb_j - \gbb)^4} + \e{(\db_j - \dbb)^4} \right)}, \label{eq:var_reduct_holder2}
	\end{align}
	where equation (\ref{eq:var_reduct_cauchy2}) follows from the Cauchy-Schwartz inequality, and equation (\ref{eq:var_reduct_holder2}) follows from H\"{o}lder's inequality.  
        Since we can bound both expectation terms by A2, we can demonstrate $\sqrt{\frac{n}{m}}(nVG - nVD) \xrightarrow{L^2} 0$ by focusing on only one of the terms.  
        Focusing on the first expectation term and returning the $\tfrac{n}{m}$, by H\"{o}lder's inequality,
	\begin{align*}
		\frac{n}{m} 4m^2 \e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}	& \leq 32 \frac{n}{m} m^2\left( \e{(\db_j - \gb_j)^4} + \e{(\dbb - \gbb)^4} \right).
	\end{align*}	
	By Lemma \ref{lem:gb_gbb_l4}, both expectation terms above converge to zero with exponential rates.  
        With a batching structure $m_l = \lfloor n_l^r \rfloor$ for some $0<r<1$, we see that $\frac{n_l}{m_l} \e{(n_l VG_l - n_l VD_l)^2} \rightarrow 0$.
\end{proof}

% END CHANGE

>>>>

\citet{Meketon1984} show that $\frac{n_l}{m_l}\var{n_l VD_l} \rightarrow \alpha \sigma^4_{\xh}$ where $\alpha = \frac{4}{3}$ for maximally overlapping batches and $\alpha = 2$ for nonoverlapping batches, resulting in the decrease of variance for the maximally overlapping batches estimator.  
\citet{Welch1987} expands this result to variable amounts of overlap, showing that $\alpha$ is twice the value of (\ref{eq:var_reduct_formula}).  
\citet{damerdji1995mean} shows that convergence of $\frac{n_l}{m_l}\var{n_l VD_l}$ to the appropriate $\alpha \sigma^4_{\xh}$---thus the mean-square consistency of $n_l VD_l$---is guaranteed for $\tfrac{1}{2} + \tfrac{1}{4+\epsilon} < r < 1$ for both nonoverlapping and maximally overlapping batches under the batching structure $m_l = \lfloor n_l^r \rfloor$ and $4+\epsilon$ finite moments assumption A2$^\prime$.  
(With A2, the range of $r$ that ensures mean-square consistency can be found by setting $\epsilon=0$.)

To prove the variance reduction results in our setting, we wish to show that 
$$
(\tfrac{n_l}{m_l}) \left( \var{n_l VG_l} - \var{n_l VD_l} \right) \rightarrow 0,
$$
for which it suffices to show $\sqrt{{n_l}/{m_l}} \left( n_l VG_l - n_l VD_l \right) \xrightarrow{L^2} 0$, where $\xrightarrow{L^2}$ denotes mean-square convergence.

@@

\begin{proof}
	For simplicity of exposition, we drop the subscript $l$ and the $\frac{n_l}{m_l}$ for now and add them in later.  
        Further, we use abbreviations $n_b = \nbl$ and $df = \left(\tfrac{n_l}{m_l} - 1\right) \left(\nbl\right)$ to denote the number of batches and the denominator term in $VG$ and $VD$ in the variably overlapping batches, respectively. Note that
	\begin{align}
		\e{(nVG - nVD)^2} & = \left( \frac{n}{df} \right)^2 \e{ \left( \sum_{j=1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right] \right)^2 } \notag \\
		& \leq \left( \frac{n}{df} \right)^2 \e{ n_b \sum_{j=1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right]^2 } \label{eq:vvar_conv_pf_1}
	\end{align}
	by the Cauchy-Schwartz inequality.  
        By examining the batch structure in Figure \ref{fig:overlap_nonint} and the definition of $\gbb$ in (\ref{eq:gbb}) one can see that the terms of the expectation are not identically distributed.  
        We will separate these into identically distributed ``center'' batches and ``fringe'' batches.  
        For example, in Figure \ref{fig:overlap_nonint}, only batch 4 is a center batch, while the others are fringe batches.  
        Notice in Figure \ref{fig:overlap_nonint}, we see $\left\lceil \gammab^{-1} \right\rceil = 3$ sets of nonoverlapping batches $\{1,4,7\}$, $\{2,5\}$, and $\{3,6\}$, each of which has two fringe batches.  
        In general, there are $\left\lceil \gammab^{-1} \right\rceil$ sets of nonoverlapping batches, each of which will have two fringe batches, and $n_b - 2\left\lceil \gammab^{-1} \right\rceil$ center batches.

	The fringe batches will come in identically distributed pairs.  
        The first and last batch would be identically distributed, as would the second and second-to-last, etc.  
        We know this because we could reverse the order of the i.i.d.\ sample and perform the batching again to get the exact same batch structure.  
        Rewriting to take this into account, letting $A_j = (\gb_j - \gbb)^2 - (\db_j - \dbb)^2$, and continuing from (\ref{eq:vvar_conv_pf_1}),
	\begin{align*}
		& \left( \frac{n}{df} \right)^2 n_b \e{ \sum_{1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right]^2 } \\
		 =  & \left( \frac{n}{df} \right)^2 n_b \e{\left[ \sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \left[ A_j \right]^2 \
		 + \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \left[ A_j \right]^2 \
		 + \sum_{j=n_b-\left\lceil \gammab^{-1} \right\rceil+1}^{n_b} \left[ A_j \right]^2 \right]\
		} \\
		 = & \left( \frac{n}{df} \right)^2 n_b \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 } + \left( \frac{n}{df} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 } \\
		 = & \left( \frac{n}{df} \right)^2 n_b \left(n_b - 2\left\lceil \gammab^{-1} \right\rceil \right) \e{ \left[ A_j \right]^2 }
		+ \left( \frac{n}{df} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 }.
	\end{align*}	
	We have $\left\lceil \gammab^{-1} \right\rceil + 1$ expectation terms, each of which has a coefficient that is less than $\left(\frac{n}{df}\right)^2n_b^2 < (2m)^2$.  
        So each expectation can be bounded by
	\begin{align}
		\left( 2m \right)^2 \e{ \left[ A_j \right]^2 } & = \left( 2m \right)^2 \e{ \left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^2 \left[(\gb_j - \gbb) + (\db_j - \dbb)\right]^2 } \notag \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}} \sqrt{\e{\left[(\gb_j - \gbb) + (\db_j - \dbb)\right]^4 }} \label{eq:var_reduct_cauchy2} \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}} \sqrt{8 \left(\e{(\gb_j - \gbb)^4} + \e{(\db_j - \dbb)^4} \right)}, \label{eq:var_reduct_holder2}
	\end{align}
	where equation (\ref{eq:var_reduct_cauchy2}) follows from the Cauchy-Schwartz inequality, and equation (\ref{eq:var_reduct_holder2}) follows from H\"{o}lder's inequality.  
        Since we can bound both expectation terms in (\ref{eq:var_reduct_holder2}) by A2, we can demonstrate $\sqrt{\frac{n}{m}}(nVG - nVD) \xrightarrow{L^2} 0$ by focusing on only one of the terms.  
        Focusing on the first expectation term in (\ref{eq:var_reduct_holder2}) and returning the $\tfrac{n}{m}$, by H\"{o}lder's inequality,
	\begin{align*}
		\frac{n}{m} 4m^2 \e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}	& \leq 32 \frac{n}{m} m^2\left( \e{(\db_j - \gb_j)^4} + \e{(\dbb - \gbb)^4} \right).
	\end{align*}	
	By Lemma \ref{lem:gb_gbb_l4}, both expectation terms above converge to zero with exponential rates.  
        With a batching structure $m_l = \lfloor n_l^r \rfloor$ for some $0<r<1$, we see that $\frac{n_l}{m_l} \e{(n_l VG_l - n_l VD_l)^2} \rightarrow 0$.
\end{proof}

Theorem \ref{thm:varvar_conv} shows conditions under which the same asymptotic variance reduction results are achieved for the OMRP optimality gap variance estimator. 
We will empirically examine the variance reduction for small sample sizes in \S \ref{sec:comp}. 
%We end our theoretical properties in the next section by examining the CIs formed using OMRP.
